---
title: "Multilevel analysis warm-up: sleepstudy"
author: "Umut Can Vural"
date: "2026-09-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Create the folder structure

```{r folder-structure}
#Create the folder structure

#dirs <- c("data", "scripts", "reports", "results", "config")
#for (d in dirs) dir.create(d, showWarnings = FALSE)

# keep empty folders visible to Git
#for (d in c("data", "results")) file.create(file.path(d, ".gitkeep"))

```

## Required packages

- **tidyverse**: Provides tools for importing, cleaning, transforming, summarising, and visualising tabular data, including `dplyr`, `readr`, `tidyr`, and `ggplot2`.

- **lme4**: Fits linear and generalized linear mixed-effects models, including random intercepts and random slopes.

- **rmarkdown**: Renders the analysis, code, figures, and written interpretation into a reproducible HTML report.

- **knitr**: Executes R code chunks and formats code, tables, figures, and output inside the R Markdown report.

- **lmerTest**: Adds p-values to lmer models

- **performance**: Provides ICC and model diagnostics.

- **performance**: Builts path from project root to files, so that the code works on any computer.

```{r packages}
library(lme4)
library(lmerTest)
library(tidyverse)
library(performance)
library(here)
```

## 01 The sleepstudy warm-up

```{r inspection}

data(sleepstudy)
glimpse(sleepstudy)

# Look at the data first: one line per person
ggplot(sleepstudy, aes(Days, Reaction, group = Subject, colour = Subject)) +
  geom_line(alpha = 0.9, linewidth = 0.8) +
  labs(title = "Reaction time over days of sleep deprivation",
       x = "Days", y = "Reaction time (ms)") +
  theme(legend.position = "none")

#before I use any predictor, how much do people differ from each other, compared with how much one person varies from measurement to measurement?
m_null <- lmer(Reaction ~ 1 + (1 | Subject), data = sleepstudy)
icc(m_null)

# Model 1: ordinary regression (ignores that measurements come from the same people)
# Reaction time gets slower per day
m0 <- lm(Reaction ~ Days, data = sleepstudy)

# Model 2: each person has their own baseline
# Everyone starts at a different level, but everyone gets slower at the same rate. 
m1 <- lmer(Reaction ~ Days + (1 | Subject), data = sleepstudy)

# Model 3: each person has their own baseline AND their own slope over time
# Everyone starts at a different level, and everyone gets slower at their own rate
m2 <- lmer(Reaction ~ Days + (Days | Subject), data = sleepstudy)

summary(m0)
summary(m1)
icc(m1)
summary(m2)
icc(m2)


# Likelihood-ratio test:
# Does m2 fit enough better to justify being more complicated?
anova(m1, m2)

```

### Model m0: ordinary regression

The average reaction time(RT) on day 0 is about **251 ms** (intercept). On average, RT gets about **10.5 ms** slower for each extra day of sleep deprivation (Days).

Individual observations scatter around the line by roughly **48 ms** (residual standard error, RSE), and the model explains about **29%** of the variance (R²).

The model assumes all 180 rows are independent, but they are 10 measurements from each of 18 people. The degrees of freedom(DoF) (**178**) are therefore too high, and the significance test is not trustworthy.

### Model m1: random intercept

The average RT on day 0 is still about **251 ms**, and the Days slope is still about **10.5 ms** per day. The standard error(St.E) of Days dropped from **1.238** to **0.804**.

People's baseline RT differ from the average by about **37 ms** (SD of the random intercept). The scatter of one person's measurements around their own line is about **31 ms** (residual SD).

Intraclass correlation coefficient (ICC) is **0.59** About **59%** of the variation in RT during these days is due to stable differences between people, and **41%** is variation within a person.

### Model m2: random intercept and random slope

The average slope is unchanged **10.5 ms**, but its standard error almost doubled **1.546**. In m1 we assumed everyone slows down at the same rate, which made the slope look more certain than it is. In m2, people are allowed to differ in their slopes, and the average slope is really estimated from 18 different people.

People's baseline RT differ from the average by about **25 ms** (SD of the random intercept). The scatter of one person's measurements around their own line is about **25.6 ms** (residual SD). People's slopes vary around the average of **10.5 ms/day** by about **5.9 ms/day.** The correlation between a person's baseline and their slope is essentially zero.

Intraclass correlation coefficient (ICC) is **0.72** About **72%** of the variation in RT during these days is due to stable differences between people, and **28%** is variation within a person. The between-person variance now depends on which day you look at, so the ICC changes over time.

### Null model and ICC

In the null model (no predictors, random intercept only), the intraclass correlation coefficient (ICC) is **0.40**. About **40%** of the total variance in reaction time is due to stable differences between people, and about **60%** is variation within a person. This justifies a multilevel approach.

### Model comparison

`anova(m1, m2)` is a likelihood-ratio test. It asks whether the more complex model (m2, random slopes) fits enough better than the simpler one (m1) to justify the extra parameters. AIC and BIC combine fit and complexity, and lower values are better.

Adding a random slope for Days improved the model substantially (AIC 1802.1 to 1763.9; χ²(2) = 42.1, p \< .001). People differ not only in their baseline reaction time but also in how quickly they slow down, so m2 is the preferred model.

## 02 Stimulate Fake Data

We simulate data from 120 older adults who complete a Digital Skills Test Measure (DSTM) at baseline (0 months), 3 months, and 6 months. Participants are assigned to either a Control group or a Training group. The Training group plays digital games daily as part of a digital-skills intervention, whereas the Control group does not.

The research question is whether the Training group improves more over time than the Control group. The main effect of interest is therefore the time-by-group interaction.

Both groups have the same average baseline score of 40 points. The Control group improves by 0.5 points per month. The Training group improves by an additional 1.0 point per month, for a total average improvement of 1.5 points per month.

Participants differ in their initial DSTM scores (sd_intercept) and individual rates of change (sd_slope). Each observed score also includes occasion-specific measurement error (sd_resid): variation around that participant’s expected score at a specific assessment.

This is a 2 × 3 mixed design:

Between-participants factor: Group (Control vs. Training)

Within-participants factor: Time (0, 3, and 6 months)

Outcome: DSTM score

The simulated random intercepts, random slopes, and residual errors are assumed to be approximately normally distributed. This assumption is part of the data-generating model and is not an application of the central limit theorem.

Finally, we fit a linear mixed-effects model using lmer() to check whether it recovers the values used to generate the data. We will later estimate power by simulating the study many times, testing the time-by-group interaction in each simulation, and calculating the proportion of significant results. We can compare power across sample sizes such N=60, N=120 and N=200, or across different intervention-effect sizes.


```{r stimulate-data}
set.seed(123456)  # makes the "random" data identical every run

# ---- Step 1. TRUE VALUES ----
# Choose the true values (averages, differences between people, noise)
# Sample and measurement occasions
n_participants <- 120
time_months <- c(0, 3, 6)

# Fixed effects
b0 <- 40          # Mean baseline DSTM score in the Control group
b_time <- 0.5     # Mean monthly DSTM change in the Control group
b_group <- 0      # Baseline difference: Training minus Control
b_time_gr <- 1.0  # Additional monthly change in the Training group

# Random effects and residual error
sd_intercept <- 8 # Between-person SD in baseline DSTM scores
sd_slope <- 0.5   # Between-person SD in monthly change rates
sd_resid <- 3     # Within-person measurement error at each occasion

# ---- 2. LEVEL 2: one row per person ----
people <- tibble(
  participant_id = sprintf("P%03d", 1:n_participants),
  intervention   = rep(c("Control", "Training"), each = n_participants / 2),
  u0 = rnorm(n_participants, 0, sd_intercept),  # this person's baseline deviation
  u1 = rnorm(n_participants, 0, sd_slope)       # this person's slope deviation
)

# ---- 3. LEVEL 1: one row per person per time point ----
sim_data <- people |>
  crossing(time_months = time_months) |>        # every person gets every time point
  mutate(
    training = as.numeric(intervention == "Training"),
    digital_skills = b0 + u0 +                                  # baseline + person deviation
      (b_time + u1) * time_months +                             # person's own slope
      b_group * training +
      b_time_gr * training * time_months +                      # intervention effect on change
      rnorm(n(), 0, sd_resid)                                   # noise
  ) |>
  select(participant_id, intervention, time_months, digital_skills) |>
  arrange(participant_id, time_months)

# ---- 4. SAVE ----
write_csv(sim_data, here("data", "simulated_digital_skills_longitudinal.csv"))
glimpse(sim_data)

#Note that u0 and u1 are the random effects: this is where "each person has their own baseline and slope" is built into the data. The random-effect part of lmer tries to recover these.

#Check: does lmer recover the truth?

#Create a second script scripts/02_check_recovery.R:

DTSM_data <- read_csv(here("data", "simulated_digital_skills_longitudinal.csv")) |>
  mutate(intervention = factor(intervention, levels = c("Control", "Training")))

# Each person has their own baseline and their own rate of change over time
# Estimate how large the effect of the intervention is on the rate of change over time
m <- lmer(digital_skills ~ time_months * intervention + (time_months | participant_id),
          data = DTSM_data)
summary(m)
```


# Fixed effects: truth vs estimate

The model recovered the simulated values well: every estimate was within about one standard error of its true value. The key interaction was significant **(𝑝=9.5×10−9)**. The Training group’s estimated total monthly improvement was **0.58+0.96=1.540** points, close to the true value of **1.5**.
The interaction’s approximate **95%** CI was **[0.65,1.26]**, which includes the true simulated effect of 1.0.

# Random effects: truth vs estimate

The model recovered the simulated random-effect values well.

The estimated variability in baseline scores, individual change rates, and measurement error was close to the true simulated values. The intercept–slope correlation of **0.12** is close to the true value of zero and is imprecise because each participant has only three measurements.

The degrees of freedom are about **118** because the group comparison is based mainly on **120** participants, not on all **360** observations. One simulation is only one random sample; power requires repeating the simulation many times and calculating how often the intervention-by-time effect is detected.


Next steps
Plot first: a spaghetti plot of individual lines coloured by group, plus the group means. This shows the story that the numbers describe.
Add confint(m) to get confidence intervals for all parameters, including the random effects.
Diagnostics: residual plot and Q-Q plot, to check the normality assumption (which holds here by design).
Model comparison: compare (1 | participant_id) with (time_months | participant_id), as in the warm-up. Note that our slope SD is only 0.5 with 3 time points, so the test may be less clear-cut than in sleepstudy.



